---
layout: post
title:  "Exploring the Colourlovers API: Getting the Data"
date:   2014-09-10 15:54:44
categories: jekyll update
---
<!--fix: change parameter from name class-->

There is a lot of great, freely available data on the web that is fascinating to explore with R.  
Companies and websites that want to expose their data to clients use a web API, where API
stands for [Application programming interface.](http://en.wikipedia.org/wiki/Application_programming_interface)  Many of the most popular APIs now have [R packages](http://cran.r-project.org/web/packages/twitteR/index.html) associated with them. These
packages help R-users retrieve data from the APIs.  But I want to show how to retrieve data from
an API of your choice
without relying on a special package for it. This will allow you to get data from a wide
range of APIs or even begin to write an R package for a specific API.

In this tutorial, I'll use the [Colourlovers API](http://www.colourlovers.com/api/) as 
an example. This API makes data from the [Colourlovers.com](http://www.colourlovers.com) website available. The website's description reads:

> COLOURlovers is a creative community where people from 
> around the world create and share colors, palettes and patterns, 
> discuss the latest trends and explore colorful articles . . . 
> All in the spirit of love.

(It turns out an [R package](http://cran.r-project.org/web/packages/colourlovers/index.html) for the Colourlovers API *was* released this August, but I first began exploring it in May.) This
API does not require authentication, which makes it ideal for a first try. 

Although I will be working without an API-specific package, I do need two R helper packages.

```{r}
library("RCurl")
library("RJSONIO")
```

First I write a function to retrieve data from the Colourlovers API.  The key helper
function 'getURL' is provided by the package **RCurl.**

```{r}
kResults <- "100"
kOrder <- "numVotes"
kSort <- "DESC"
kFormat <- "json"
kKeyword <- "ombre"

GetCLURL <- function(parameters, values, api.class){
  if(length(parameters) != length(values)){
    stop("The length of the parameters and the 
         length of the values must be the same.")
  }
  cl.URL.palettes <-
    paste("http://www.colourlovers.com/api/", api.class, "?", sep="")
  cl.URL.options <- 
    paste(paste(parameters, values, sep="="), collapse="&")
  cl.URL <- paste(cl.URL.palettes, cl.URL.options, sep="")
  cl.json <- getURL(cl.URL)
  return(cl.json)
}
```
Now I call the function I have written.  I'm interested in color palettes
that contain the keyword 'ombre.'  I'm requesting that data be returned to me
in JavaScript Object Notation, or [JSON](http://en.wikipedia.org/wiki/JSON).

```{r}
my.parameters <- c("numResults", "orderCol", "keywords", "sortBy", "format")
my.values <- c(kResults, kOrder, kKeyword, kSort, kFormat)
ombre.json <- GetCLURL(my.parameters, my.values, "palettes")
class(ombre.json)
length(ombre.json)
head(ombre.json)
```
The return value is simply a very long string in JSON format.  I can convert the JSON string into an R object using the function 'fromJSON' from the R package **RJSONIO.**

```{r}
ombre.processed <- fromJSON(ombre.json)
class(ombre.processed)
length(ombre.processed)
class(ombre.processed)[[1]]
```

The R object returned is a list of lists. Each sublist represents a color palette.

```{r}
ombre.processed[[1]]
```

Moreover, the sublist elements are of different types.

```{r}
class(ombre.processed[[1]][[1]])
class(ombre.processed[[1]][[2]])
```
Now I save the list of lists for later manipulation.

```{r}
save(ombre.processed, file="partI.RData")
```

Next up: *Exploring the Colourlovers API: Data Cleaning and Transformation.*